Rates and delay times of type Ia supernovae in the Dark Energy Survey

### 2021-02-17 draft.tex

Type Ia supernovae (SNe Ia) are explosions of white dwarves (WDs) at masses close to the Chandrasekhar limit of 1.44 $M_{\odot}$. The small observed dispersion in their peak brightnesses can be reduced further through empirical relations between brightness and light curve properties, such as decline rate (stretch) or optical colour. These properties have led SNe Ia to be used extensively by cosmologists to measure relative distances in the Universe.

Despite reducing dispersions to $\sim 0.1$ mag in samples with over one thousand SNe \citep{Scolnic2018}, the exact nature of the progenitor scenarios for SNe Ia is yet to be confirmed. While it is accepted that SNe Ia are caused by accretion of matter onto a WD from a companion star, multiple possible scenarios exist for the nature of that companion star and the search for this has formed a significant field of research (see \citealt{Maoz2014} for a review). The leading models involve a main sequence (MS) star in the single-degenerate (SD) scenario \citep{Whelan1973,Nomoto1982}, a secondary white dwarf in the double-degenerate (DD) scenario \citep{Tutukov1976,Iben1984,Webbink1984}, while more exotic models such as the core-degenerate (CD) scenario invoke white dwarves merging with the cores of massive stars \citep{Soker2019}.

The progenitors of a number of core-collapse SNe (CCSNe) have been identified in high-resolution pre-explosion images, allowing detailed analysis of the latter stages of massive star evolution. However, this approach has not yet been successful in identifying a progenitor of a SN Ia. Conversely, it has been possible to search for surviving remnants of the binary system -- main sequence stars or white dwarves left behind at the SN location or kicked \citep{Shen2018} -- displaying tentative evidence for a DD scenario. As yet, there is no conclusive evidence ruling out any progenitor channel from being a significant contribution to the total population of SNe Ia.

Another method of inferring progenitor channels is the study of the SN Ia delay time distribution (DTD). The DTD can be interpreted as the rate at which SNe Ia occur as a function of the delay time $\tau$ since an episode of star-formation, and thus carries characteristic signatures of the progenitor channel (see \citealt{Wang2012} for a review). While the SD scenario is able to produce a broad range of functional forms of the DTD, the DD scenario predicts a power law of the form $\tau^{\beta}$, with $\beta \sim -1$ \citep[e.g][]{Ruiter2009,Mennekens2010}.

A power law DTD implies that most SNe Ia explode soon after star formation, but also that there is a non-neglible fraction that explode on the order of Gyrs after the star formation episode. The DTD is therefore non-trivial to measure and various techniques have been developed to infer it indirectly. One such approach is to measure global properties of SN Ia host galaxies. Simple examples of such analyses are comparisons of the rates of SNe Ia in hosts that have been split by some observational property, such as morphology or colour, and have shown that SNe Ia rate per unit stellar mass (SNuM) is significantly larger in late-type spiral galaxies as well as in bluer galaxies \citep[e.g.][]{Mannucci2005}. These observations heavily imply that SNe Ia rates are dependent on recent or ongoing star-formation, and thus that the majority of SNe Ia explode after short delay times. Conversely, the SNuM in E/S0 and red galaxies is non-negligible, suggesting that there is a secondary, much older component to the DTD \citep{Sullivan2006,Smith2012}. The DTD is thus approximated by a two-component ("$A + B$") model, where $A$ and $B$ are the normalizations of the DTD for "prompt" (i.e. proportional to instantaneous SFR) and "tardy" (i.e. proportional to overall stellar mass) SNe, respectively.

@@

It is also possible to estimate the SFH of an individual galaxy through the modelling of its stellar populations via spectral energy distribution (SED) fitting. Works such as \citet{Maoz2011}, \citet{Maoz2012} and \citet{Graur2013} estimated the DTD by measuring SFHs for a sample of field galaxies and comparing them to the number of SNe detected in each galaxy (the SFHR method), and have led to results that suggest a DTD power law with $\beta$ consistent with $-1$ \citep[e.g.][]{Maoz2011,Graur2013}. \citet{Childress2014} showed that the $A+B$ parameterisation arises as a direct consequence of the combination of a power-law DTD with the SFHs of galaxies. The recent advances of integral field spectroscopy (IFS) allows for an extension of the SFHR method, by reconstructing the SFH for hundreds or thousands of local regions of each SN Ia host galaxy. Using this method, \citet{Castrillo2020} find a power-law slope of $-1.1\pm0.3$, while \citet{Chen2021} find $-1.4\pm0.3$.

@@

We introduce the SN host and field galaxy samples in Section \ref{sec:data} and describe our handling of incompleteness in Section \ref{sec:incompleteness}. We present the SN Ia rate per galaxy in Section \ref{sec:rates}. We outline the modelling of star-formation histories and measurement of the DTD in Section \ref{sec:model} before concluding in Section \ref{sec:conclusion}.
Where relevant, we assume a spatially flat $\Lambda$CDM cosmology with $\Omega_m = 0.3$ and $H_0 = 70 \mathrm{ km s}^{-1}\mathrm{Mpc}^{-1}$. Unless otherwise stated, we assume Gaussian uncertainties quoted at the $1\sigma$ level. Magnitudes are quoted in the AB system \citep{Oke1983}.

@@

\subsection{Splitting by SN stretch \label{subsec:split_x1}}

The SN Ia rate as a function of stellar mass SNe split by stretch is shown in Fig. \ref{fig:rate_fitted_split_x1}. The evolution of the rate with stellar mass is significantly different for the two sub-samples: high-stretch SNe dominate the rate at low stellar mass but tail off in the higher mass galaxies, while low-stretch SNe are subdominant in low mass galaxies but display a much steeper dependence on stellar mass and make up the vast majority of SNe in high-mass hosts. These measurements reflect the observed correlations of $x_1$ with stellar mass and sSFR. The corresponding DTD fit results in a steep power-law for high-stretch SNe indicative of a population of predominantly prompt SNe, and a much shallower decay slope for low-stretch SNe representing a much more delayed population. Such a distinction has been predicted by \citet{Rigault2013,Childress2014,Rigault2018,Nicolas2020} based on the intrinsic $x_1$ distribution and its observed relationship with stellar mass and age.

From the differing DTDs that describe SNe with low and high stretch values we infer that there are either multiple channels through which SN Ia explode which occur on differing timescales, or that the explosion mechanism evolves with progenitor age. Two scenarios, that are not necessarily mutually exclusive, that could lead to different DTD decay rates are as follows. 
The first such scenario is to assume that all SNe Ia come from the same initial population of stars, and thus that the progenitors of both low and high stretch SNe form at the same time. It is then necessary to invoke models in which the WD progenitors of low-stretch SNe evolve on longer timescales than those of high-stretch SNe, for example due to a different initial binary separation distribution or accretion rate.

Alternatively, low- and high-stretch SNe could form via identical evolutionary channels but with a different onset time since the episode of star-formation. The predominantly high-stretch "prompt" SNe would begin exploding as soon as the WDs are formed (nominally 40 Myr) whereas the mainly low-stretch "tardy" SNe would only begin exploding after an extended period of time of the order 1 Gyr. The rates of both populations would then both follow the fiducial $t^{-1}$ distribution beginning at this onset time.
Such a scenario has been explored theoretically, primarily in the sub-Chandrasekhar mass (sub-$M_{\mathrm{Ch}}$) paradigm \citep[e.g.][]{Sim2010,Blondin2017,Shen2017}, usually associated with various forms of the DD scenario. Sub-$M_{\mathrm{Ch}}$ SNe Ia typically involve the merger of two WDs where the SN luminosity is correlated with the mass of the primary WD; a correlation between primary WD mass and age would then lead to the different rates, and apparent DTDs, of low and high-stretch SNe observed in this work.

@@

The SN Ia rates generated by the posterior predictive checks are shown in Fig. \ref{fig:rate_fitted_split_x1_joint}. The best fitting values are $t_1 = 0.65 \pm 0.1$~Gyr and $t_2 = 0.85 \pm 0.15$~Gyr, corresponding to a relatively sharp transition between the two populations.

\subsubsection{The late end of the DTD \label{subsubsec:subtypes}}

>>>>

Type Ia supernovae (SNe Ia) are explosions of white dwarf stars (WDs) at masses close to the Chandrasekhar limit of 1.44 $M_{\odot}$. The small observed dispersion in their peak brightnesses can be reduced further through empirical relations between brightness and light curve properties, such as decline rate (stretch) or optical colour \citep{Philips1993,Tripp1998}. These properties have led SNe Ia to be used extensively by cosmologists to measure relative distances in the Universe \citep{Perlmutter1998,Riess1998}

Despite using these relations to correct SN Ia peak brightnesses to within a dispersion of $\sim 0.1$ mag in samples with over one thousand SNe \citep{Scolnic2018}, the exact nature of the progenitor scenarios for SNe Ia is yet to be confirmed. While it is accepted that SNe Ia are caused by mass transfer onto a WD from a companion star, multiple possible scenarios exist for the nature of that companion star (see \citealt{Maoz2014} for a review). The leading models involve a main sequence (MS) star in the single-degenerate (SD) scenario \citep{Whelan1973,Nomoto1982}, a secondary white dwarf in the double-degenerate (DD) scenario \citep{Tutukov1976,Iben1984,Webbink1984}, while more exotic models such as the core-degenerate (CD) scenario invoke white dwarves merging with the cores of massive stars \citep{Soker2019}.

The progenitors of a number of core-collapse SNe (CCSNe) have been identified in high-resolution pre-explosion images, allowing detailed analysis of the latter stages of massive star evolution. However, this approach has not yet been successful in identifying a progenitor of a SN Ia. Conversely, it has been possible to search for surviving remnants of the binary system -- main sequence stars or white dwarves left behind at the SN location or kicked \citep{Shen2018} -- displaying tentative evidence for a DD scenario. Without an observation of a SN Ia progenitor system or remnant companion, there is as yet no direct evidence ruling out any progenitor channel from being a significant contribution to the total population of SNe Ia.

Another method of inferring progenitor channels is the study of the SN Ia delay time distribution (DTD). The DTD describes the rate at which SNe Ia occur as a function of the delay time $\tau$ since an episode of star-formation, and thus carries characteristic signatures of the progenitor channel (see \citealt{Wang2012} for a review). While the SD scenario is able to produce a broad range of functional forms of the DTD, most variants of the DD scenario predict a power-law of the form $\tau^{\beta}$, with $\beta \sim -1$ \citep[e.g][]{Ruiter2009,Mennekens2010}.

A power law DTD implies that most SNe Ia explode soon after star formation, but also that there is a non-neglible fraction that explode on the order of Gyrs after the star formation episode. The DTD is therefore non-trivial to measure and various techniques have been developed to infer it indirectly. One such approach is to measure global properties of SN Ia host galaxies. Simple examples of such analyses are comparisons of the rates of SNe Ia in hosts that have been split by some observational property, such as morphology or colour, and have shown that SN Ia rate per unit stellar mass (SNuM) is significantly larger in late-type spiral galaxies as well as in bluer galaxies \citep[e.g.][]{Mannucci2005}. These observations heavily imply that SNe Ia rates are dependent on recent or ongoing star-formation, and thus that the majority of SNe Ia explode after short delay times. Conversely, the SNuM in E/S0 and red galaxies is non-negligible, suggesting that there is a secondary, much older component to the DTD \citep{Sullivan2006,Smith2012}. The DTD is thus approximated by a two-component ("$A + B$") model, where $A$ and $B$ are the normalizations of the DTD for "prompt" (i.e. proportional to instantaneous SFR) and "tardy" (i.e. proportional to overall stellar mass) SNe, respectively.

@@

Instead of comparing \textit{volumetric} rates to the \textit{cosmic} SFH, it is also possible to estimate the SFH of individual galaxies through the modelling of their stellar populations via spectral energy distribution (SED) fitting. Works such as \citet{Maoz2011}, \citet{Maoz2012} and \citet{Graur2013} estimated the DTD by measuring SFHs for a sample of field galaxies and comparing them to the number of SNe detected in each galaxy (the SFHR method), and have led to results that suggest a DTD power law with $\beta$ consistent with $-1$ \citep[e.g.][]{Maoz2011,Graur2013}. \citet{Childress2014} showed that the $A+B$ parameterisation arises as a direct consequence of the combination of a power-law DTD with the average SFHs of galaxies. The recent advances of integral field spectroscopy (IFS) allows for an extension of the SFHR method, by reconstructing the SFH for hundreds or thousands of local regions of each SN Ia host galaxy. Using this method, \citet{Castrillo2020} find a power-law slope of $-1.1\pm0.3$, while \citet{Chen2021} find $-1.4\pm0.3$.

@@

We introduce the SN host and field galaxy samples in Section \ref{sec:data} and describe our handling of incompleteness in Section \ref{sec:incompleteness}. We present the SN Ia rate per galaxy in Section \ref{sec:rates}. We outline the modelling of star-formation histories and measurement of the DTD in Section \ref{sec:model}. In Section \ref{sec:split_x1_c} we investigate how the DTD differs among sub-populations of SNe with different light curve characteristics. We conclude in Section \ref{sec:conclusion}.
Where relevant, we assume a spatially flat $\Lambda$CDM cosmology with $\Omega_m = 0.3$ and $H_0 = 70 \mathrm{ km s}^{-1}\mathrm{Mpc}^{-1}$. Unless otherwise stated, we assume Gaussian measurement uncertainties quoted at the $1\sigma$ level, and we quote $1\sigma$ credible intervals on derived parameters. Magnitudes are quoted in the AB system \citep{Oke1983}.

@@

\subsection{Splitting by SN stretch \label{subsec:split_x1}}
\begin{figure}
    \centering
    \includegraphics[width=.5\textwidth]{figs/rate_vs_mass_DTD_fit_beta_norm_Qerf1.1_split_x1_fitjoint.png}
    \caption{As per Fig. \ref{fig:rate_fitted_split_x1} but with the data fit by a two-component DTD.%The blue lines indicate the best fitting model based on a $t^{\beta}$ power law DTD convolved with SFHs based on the \citet{Childress2014} model}
    \label{fig:rate_fitted_split_x1_2c_DTD}}
\end{figure}
The SN Ia rate as a function of stellar mass SNe split by stretch is shown in Fig. \ref{fig:rate_fitted_split_x1}. The evolution of the rate with stellar mass is significantly different for the two sub-samples: high-stretch SNe dominate the rate at low stellar mass but tail off in the higher mass galaxies, while low-stretch SNe are subdominant in low mass galaxies but display a much steeper dependence on stellar mass and make up the vast majority of SNe in high-mass hosts. These measurements reflect the observed correlations of $x_1$ with stellar mass and sSFR. The corresponding DTD fit results in a steep power-law $\beta = -1.68 \pm 0.15$ for high-stretch SNe indicative of a population of predominantly prompt SNe, and a much shallower decay slope ($\beta = -0.79 \pm 0.08$) for low-stretch SNe representing a much more delayed population. Such a distinction between SN stretch and progenitor age has been seen by \citet{Rigault2013,Childress2014,Rigault2018,Nicolas2020}; here we have shown it purely by measuring the rate against stellar mass.

From the differing DTDs that describe SNe with low and high stretch values we infer that there are either multiple channels through which SN Ia explode which occur on differing timescales, or that the explosion mechanism evolves with progenitor age. Two scenarios, that are not necessarily mutually exclusive, that could lead to different DTD decay rates are as follows. 
The first such scenario is to assume that all SNe Ia come from the same initial population of stars, and thus that the progenitors of both low and high stretch SNe form at the same time. It is then necessary to invoke models in which the WD progenitors of low-stretch SNe evolve on longer timescales than those of high-stretch SNe, for example due to a different initial binary separation distribution or accretion rate. In this case, the DTD slopes would indeed take on the different values measured here

Alternatively, low- and high-stretch SNe could form via identical evolutionary channels but with a different onset time since the episode of star-formation. The predominantly high-stretch "prompt" SNe would begin exploding as soon as the WDs are formed (nominally 40 Myr) whereas the mainly low-stretch "tardy" SNe would only begin exploding after an extended period of time of the order 1 Gyr. The rates of both populations would both follow the fiducial $t^{-1}$ distribution.
Such a scenario has been explored theoretically, primarily in the sub-Chandrasekhar mass (sub-$M_{\mathrm{Ch}}$) paradigm \citep[e.g.][]{Sim2010,Blondin2017,Shen2017}, usually associated with various forms of the DD scenario. Sub-$M_{\mathrm{Ch}}$ SNe Ia typically involve the merger of two WDs where the SN luminosity is correlated with the mass of the primary WD; a correlation between primary WD mass and age would then lead to the different rates, and apparent DTDs, of low and high-stretch SNe observed in this work. Such an evolution of light curve properties with age has recently been seen across the full range of "normal" SNe Ia in 1-dimensional simulations by \citet{Shen2021} and is a promising avenue for further investigation.

@@

The SN Ia rates generated by the posterior predictive checks are shown in Fig. \ref{fig:rate_fitted_split_x1_2c_DTD}. The best fitting values are $t_1 = 0.65 \pm 0.1$~Gyr and $t_2 = 0.85 \pm 0.15$~Gyr, corresponding to a relatively sharp transition between the two populations. In reality, we might expect that the transition is not so sharp, but a gradual correlation of delay time with the average stretch of SNe caused by the age vs WD-mass distribution, as predicted by \citet{Shen2021}. However we note that the intrinsic stretch distribution may not be a simple Gaussian, but appears to be bimodal \citep{Scolnic2016,Popovic2021}. The correlation of progenitor age and WD mass does not easily reproduce such a bimodal distribution, and we encourage further study of the explosion mechanism in order to reconcile this difference.
\subsubsection{The late end of the DTD \label{subsubsec:subtypes}}
